---
title: "basics"
author: "Raven McKnight"
date: "10/15/2019"
output: 
  html_document:
    theme: paper
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
packages <- c('data.table', 'rstan', 'ggplot2', 'bayesplot', 'tigris', 'sf', 'dplyr', 
              'CARBayes', 'spdep', 'corrplot', 'reshape2', 'scales', 'leaps')

miss_pkgs <- packages[!packages %in% installed.packages()[,1]]

if(length(miss_pkgs) > 0){
  install.packages(miss_pkgs)
}

invisible(lapply(packages, library, character.only = TRUE))

rm(miss_pkgs, packages)

mod_dat <- readRDS('/Users/raven/Documents/honors/honors-work/data/ag_2017_scaled_mod.RDS')
setDT(mod_dat)

options(tigris_class = 'sf')
counties <- c('Anoka', 'Carver', 'Dakota', 'Hennepin', 'Ramsey', 'Scott', 'Washington')
bgs <- block_groups('MN', counties, 2016)

spatdat <- left_join(bgs, mod_dat, on = 'GEOID')
```

# Variables

The variables we're working with now are logged and scaled to have mean 0 variance 1. Let's just look at them. 

```{r}
df <-melt(mod_dat[, -c("GEOID")], id=c("avg_act_per_capita")) 

ggplot(mod_dat, aes(x=avg_act_per_capita)) +
  geom_histogram() +
  labs(title = "response variable") +
  theme_minimal()

ggplot(df, aes(x=value)) +
  geom_histogram() +
  facet_wrap(~variable, scales = "free") +
  labs(title = "All our variables") +
  theme_minimal()

ggplot(df, aes(x=value, y=avg_act_per_capita)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Visual correlation?") +
  theme_minimal()
```

It's a little hard to see above but I think all the general trends make sense. 

# Correlation
```{r}
cor_dat <- mod_dat[!is.na(estimate_median_hh_income), -c('GEOID', 'sqkm', 'estimate_tot_pop', 'tot_jobs', 
                                                         'w_total_jobs_here')]
cor_matrix <- round(cor(x=cor_dat), 2)
cor_melt <- melt(cor_matrix)

ggplot(cor_melt, aes(x=Var1, y=Var2, fill=abs(value))) + 
    geom_tile() + 
    theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Okay I do think there's some interesting stuff going on with the percent jobs less 40. Lot's of the correlations that stick out are kinda silly.

```{r}
ggplot(cor_melt, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_fill_gradientn(colors = c(low = muted("red"), mid = "white", high = muted("blue")))
  
```



# Spatial clustering?

I expect the Moran's I to be relatively low because it tests for *clustering* and our spatial structure here is more linear. This will also drop about 1000 rows with nas but that's okay for now. 

```{r}
W <- poly2nb(spatdat)
W_list <- nb2listw(W, style = "B")

moran.mc(x = na.omit(spatdat$avg_act_per_capita), listw = W_list, nsim = 1000, zero.policy = TRUE)
```

As expected, moderate spatial clustering!

# Linear models

```{r}
mod1 <- lm(avg_act_per_capita ~ emp_density + perc_rent + perc_no_veh, data = mod_dat)
summary(mod1)
```

P-values will always be low like this until we incorporate **level of service** predictors. The most basic is number of trips per day, although this is super flawed!!

```{r}
mod2 <- lm(avg_act_per_capita ~ emp_density + perc_rent + perc_no_veh + avg_trips, data = mod_dat)
summary(mod2)
```

Nice! For fun, all in:

```{r}
mod3 <- lm(avg_act_per_capita ~ ., data = mod_dat[, -c('GEOID')])
summary(mod3)
```

Interesting: the only things that are significant at any level are:

- median age
- total pop
- wac jobs less than 40k
- wac total jobs
- employment density 
- percent no vehicle
- percent rent
- number of trips

Employment models. 

```{r}
mod4 <- lm(avg_act_per_capita ~ emp_density + avg_trips, mod_dat)
summary(mod4)
```

Just rac
```{r}
mod5 <- lm(avg_act_per_capita ~ emp_density + perc_jobs_white + perc_jobs_men + perc_jobs_no_college + perc_jobs_less40 + perc_jobs_age_less30 + avg_trips, mod_dat)
summary(mod5)
```
Okay interesting that age of worker overpowers wage. 

Same mod but with wac:
```{r}
mod6 <- lm(avg_act_per_capita ~ emp_density + w_perc_jobs_white + w_perc_jobs_men + w_perc_jobs_no_college + w_perc_jobs_less40 + w_perc_jobs_age_less30 + avg_trips, mod_dat)
summary(mod6)
```

Okay, interesting.. what does the difference between wac & rac mean??

```{r}
mod7 <- lm(avg_act_per_capita ~ emp_density + avg_trips + perc_jobs_less40 + w_perc_jobs_less40
           + perc_jobs_age_less30 + w_perc_jobs_age_less30, mod_dat)
summary(mod7)
```

Seems like resident area characteristics are slightly more influential. Overall this has been interesting but I'm going to try some stepwise selection I think. 

# Best Subsets

```{r}
small_dat <-  mod_dat[, -c("perc_bach", "tot_jobs", "w_total_jobs_here", "sqkm", "GEOID", "estimate_tot_pop")]
best_subsets <- regsubsets(avg_act_per_capita ~ ., small_dat, nvmax = 21)

best_sum <- summary(best_subsets)

pd <- data.frame(size = 1:20, adjr2 = best_sum$adjr2)
ggplot(pd, aes(x = size, y = adjr2)) + 
  geom_point() + 
  geom_line() + 
  labs(x="subset size", y="Adjusted R-squared") + 
  theme_minimal()
```

Seems like 12 is the absolute max adjusted $R^2$. 

The 12 predictors are

- avg_trips
- estimate_median_hh_income
- perc_only_white
- perc_hs
- perc_no_veh
- perc_english_only
- emp_density
- perc_jobs_age_less30
- w_perc_jobs_white
- w_perc_jobs_men (surprising?? which sign is the coefficient??)
- w_perc_jobs_no_college
- w_perc_jobs_less40


Best subset lm:
```{r}
bss <- lm(avg_act_per_capita ~ avg_trips + estimate_median_hh_income + perc_only_white + perc_hs + perc_no_veh + perc_english_only
          + emp_density + perc_jobs_age_less30 + w_perc_jobs_white + w_perc_jobs_men + w_perc_jobs_no_college + w_perc_jobs_less40, mod_dat)

summary(bss)
```

w_perc_men coeff is positive -- so men are slightly more likely to commute via transit? Suprising: positive coef on perc_english_only, negative coef on w_perc_jobs_no_college. 

Might move forward with the best subset of 6 for now
```{r}
bss6 <- lm(avg_act_per_capita ~ avg_trips + estimate_median_hh_income + perc_hs + emp_density + perc_jobs_age_less30 + w_perc_jobs_white, mod_dat)
summary(bss6)
```


# Spatial residuals

Going to run with bss6 for now. 

Let's check the spatial dependence in the residuals. 


```{r}
resids <- residuals(bss6)

resid_dat <- mod_dat[!is.na(avg_act_per_capita) & !is.na(estimate_median_hh_income) & !is.na(emp_density)]
resid_dat$resids <- resids

resid_dat <- left_join(bgs, resid_dat, by = 'GEOID')
Wr <- poly2nb(resid_dat)
Wr_list <- nb2listw(Wr, style = "B")

moran.mc(x = na.omit(resid_dat$resids), listw = Wr_list, nsim = 1000, zero.policy = TRUE)
```

Okay so that handled a decent about of the clustering -- we know there's still significant "linear clustering." We might also expect that some of the remaining variation could be explained by spatial variation?


- don't need to standardize y

