---
title: "Most Basic Models"
author: "Raven McKnight"
date: "9/30/2019"
output: 
  html_document:
    theme: paper
    code_folding: hide
---

```{r setup, warning = FALSE, message = FALSE, results = 'hide'}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
packages <- c('data.table', 'rstan', 'ggplot2', 'bayesplot', 'tigris', 'sf', 'dplyr')

miss_pkgs <- packages[!packages %in% installed.packages()[,1]]

if(length(miss_pkgs) > 0){
  install.packages(miss_pkgs)
}

invisible(lapply(packages, library, character.only = TRUE))

rm(miss_pkgs, packages)
rstan_options(auto_write = TRUE)

options(tigris_class = 'sf')
counties <- c('Anoka', 'Carver', 'Dakota', 'Hennepin', 'Ramsey', 'Scott', 'Washington')
bgs <- block_groups('MN', counties, year = 2016)
bgs <- st_transform(bgs, 32615) # let's go straight to mapping
```

This document will explore some of **the most** basic models for ridership. 

## Variables

### Response

I'll start with the most basic, most aggregated version of a ridership model. For now, the response variable will be **average 2017 daily boardings + alightings/total square kilometers by Census block group**. I don't currently have a reason for using square kilometers over other measures of area, other than that the Census distributes land area in meters. Here is the distribution of that response variable. 
```{r}
# not the ideal way to read in data
apc <- readRDS('/Users/raven/Documents/honors/honors-work/data/mt-data/apc-bg-ag-sum.RDS')
setDT(apc)

# restrict to 2017
apc <- apc[date_key %like% 2017]

# get averages
apc <- apc[, .(avg_activity = mean(ag_board + ag_alight)), by = GEOID]

# join to block groups
apc <- left_join(bgs, apc, by = 'GEOID')

# get response var
apc <- apc %>%
  mutate(land_km = ALAND/1000000, 
         avg_activity_dens = avg_activity/land_km)

ggplot(apc, aes(x=avg_activity_dens)) +
  geom_histogram() + 
  theme_minimal() +
  labs(x = 'activity', y = 'count', title = 'Distribution of average daily activity per square kilometer, 2017') +
  theme(plot.title = element_text(hjust = 0.5))
```

There are a few huge outliers, mostly at transit centers in the downtowns. Log transformed response variable: 

```{r}
ggplot(apc, aes(x=log(avg_activity_dens))) +
  geom_histogram() + 
  theme_minimal() +
  labs(x = 'activity', y = 'count', title = 'Log distribution of average daily activity per square kilometer, 2017') +
  theme(plot.title = element_text(hjust = 0.5))
```

Generally, ridership always needs to be transformed because those outliers will always be there. 

### Covariates

I'm going to start with some basic demographic predictors. Strangely, Met Council never used any demographics (other than population density) as predictors. Intuition & experience at Metro Transit definitely suggest that basic indicators like household income & language spoken at home should have substantial effects. 

```{r}
acs <- readRDS('/Users/raven/Documents/honors/honors-work/data/Covariates/basic_acs.RDS')
setDT(acs)

# restrict to 2017
acs <- acs[year == 3]

ggplot(acs[variable != 'speak_only_english' & variable != 'tot_no_veh'], aes(x=estimate)) +
  geom_histogram() + 
  facet_wrap(~variable) +
  theme_minimal() +
  labs(x = 'activity', y = 'count', title = 'Distribution of covariates, 2017') +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(acs[variable != 'speak_only_english' & variable != 'tot_no_veh'], aes(x=log(estimate))) +
  geom_histogram() + 
  facet_wrap(~variable) +
  theme_minimal() +
  labs(x = 'activity', y = 'count', title = 'Log distribution of covariates, 2017') +
  theme(plot.title = element_text(hjust = 0.5))
```

## Ridership ~ Population

## Ridership ~ Household income

## Ridership ~ Vehicle Deficit